A community-maintained standard library of population genetic simulations

### 2019-12-18 popSimManu.tex
some reworking of Discussion

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.png}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairwayplot},
\MSMC with $n=2$ and $n=8$. Data were generated by simulating
replicate human genomes under the \cite{ragsdale2019models} model and using the
HapMapII genetic map~\citep{international2007second}. From top to bottom we show estimates for each
of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:n_t_ragsdale}
\end{center}
\end{figure}

@@

We have built \stdpopsim, and the PopSim Consortium, to be an open, community-developed
project, meaning that it has excellent potential for future growth.
Thus far we have implemented genome representations and genetic maps for the three most
common study systems in computational genetics: humans, \emph{Drosophila}, and \emph{Arabidopsis}.
The standard operating procedures that we have outlined in our detailed documentation
means that new researchers, who are interested in adding their favorite organism/genome
are free to do so with very few barriers to entry (see Methods).
Further, new demographic models can be added
to each of the organisms in the catalog through similarly well documented steps.
A long-term goal then is that through a mechanism such as \stdpopsim, the community
can determine (in an open way) what is the correct implementation for a published demographic model,
and in turn, what are the models in which the community is most interested.

We have shown in the results above how \stdpopsim can be used for direct comparisons
of inferential methods on a common set of simulations. While we have not made a concerted
effort at benchmarking methods yet, we believe that this would be an important goal for
the PopSim Consortium and population genetics more broadly. The analysis workflows that
we have released as part of \stdpopsim should provide a cookbook for the community to
dive in to more rigorous benchmarking than we have done thus far.

In the use case comparisons we have performed,
we found that the methods tested \emph{do} generally perform well on human genomes,
up to, for instance, the usual factor-of-two differences in inferred population sizes.
This will not be surprising to people familiar with the landscape of
demographic inference methods in population genetics,
but the nature and scale of uncertainty may be not clear to people new to the field.
The ability for \stdpopsim to generate data with and without significant features, such
as the genetic map or population size change (e.g., Figure \ref{fig:n_t_HomSap_constant}), allows
us to ask directly about the various failure modes of popular methods.
Moreover the comparison of methods across the various genome organizations, genetic maps,
and demographic histories of different organisms, provides valuable information
about how methods might perform on non-human systems.
In our initial comparison we find that inference of population histories from simulated \emph{Drosophila}
or \emph{Arabidopsis} genomes is a fair bit more difficult from SMC based methods
than in the human genome. Likely this is at least partly a consequence of the reduced physical and genetic
map lengths of these genomes, although we will leave further investigation of this to the future.
Finally, comparison of results across methods or simulation runs
gives a strong estimate of inference uncertainty, analogous to parametric
bootstrapping---although different methods have different strengths and weaknesses,
strong conclusions should probably not rest entirely on features inferred by only one method.

The broad brush comparisons we have done, while far from exhaustive, are informative,
and highlight the utility of easy comparison of methods on the same footing.
The tests here were under best-case scenarios---simulations of discrete,
randomly mating populations that fit the assumptions of the methods nearly
exactly---and yet different methods, as well as applications of the same method
to different simulations of the same model,
produced somewhat different results. Future efforts will extend the resource described here to
encompass additional population genetic tasks, such as detecting natural selection, inferring
recombination rates, etc, in hopes of producing a general-purpose resource for fair and
standardized evaluation of the ever-increasing arsenal of computational population genetic
inference tools.

@@

\pagebreak
\beginsupplement
\section*{Supplemental Figures}
\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.png}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Estimates of population
size over time ($N(t)$) inferred using 4 different methods, \smcpp, \stairwayplot,
\MSMC with $n=2$ and $n=8$. Data were generated by simulating
replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
inferred in \cite{international2007second}. From top to bottom we show estimates for each
of the three populations in the model: YRI, CEU, and CHB. In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:n_t_gutenkunst}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_constant.png}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
size over time ($N(t)$) inferred using 4 different methods, \smcpp, \texttt{stairwayplot},
\MSMC with $n=2$ and $n=8$. Data were generated by simulating
replicate human genomes under a constant sized population model with $N=10^4$ and using the
HapMapII genetic map~\citep{international2007second}. In black we show the true population size history
of the model.}
\label{fig:n_t_HomSap_constant}
\end{center}
\end{figure}

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Ragsdale.pdf}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
size over time ($N(t)$) inferred using 4 different methods: \smcpp, \texttt{stairwayplot},
\MSMC with $n=2$ and $n=8$. Data were generated by simulating
replicate human genomes under the \cite{ragsdale2019models} model and using the
HapMapII genetic map~\citep{international2007second}. From top to bottom we show estimates for each
of the three populations in the model (YRI, CEU, and CHB). In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:n_t_ragsdale}
\end{center}
\end{figure}

@@

We have illustrated in this paper how \stdpopsim can be used for direct
comparisons of inferential methods on a common set of simulations. Our
benchmarking comparisons have been limited so far, but nevertheless, they
do reveal some broad generalizations about relative performance. For
example, at the task of estimating $N(t)$ trajectories for single human
populations, we find that the sequence-based methods (\MSMC and \smcpp)
perform somewhat better overall---at least for mid-range values of
$t$---than the site frequency spectrum-based method (\stairwayplot)
(Figure~\ref{fig:n_t_ragsdale} and Figure~\ref{fig:n_t_gutenkunst}), which
tends to over-estimate oscillations in size.  By contrast, in the case of
\emph{D.~melanogaster}, where linkage disequilibrium is considerably
reduced, \stairwayplot outperforms the sequence-based methods
(Figure~\ref{fig:n_t_sheehan}). In the two-population setting for human
(Figure~\ref{fig:IM_popn_human}), most methods do reasonably well at
reconstructing the YRI history, but struggle with the more complex CEU
history, in large part because of the restriction of constant sizes per
population.  An exception is \smcpp, which does not have the same
restrictions on its inferred history, and as a result does somewhat better
with the CEU history but tends to overfit the YRI history.  The results for
the two-population {\em Drosophila} case (Figure~\ref{fig:two_popn_fly})
are generally similar.

Altogether, these preliminary experiments highlight
the utility of \stdpopsim for comparing a variety of inference methods on
the same footing, under a variety of different demographic models.
In addition, the ability of \stdpopsim to generate data with and without significant features, such
as the genetic map or population size change (e.g., Figure \ref{fig:n_t_HomSap_constant}), allows
investigation of the failure modes of popular methods.
Moreover the comparison of methods across the various genome organizations, genetic maps,
and demographic histories of different organisms, provides valuable information
about how methods might perform on non-human systems.
Finally, comparison of results across methods or simulation runs
provides an estimate of inference uncertainty, analogous to parametric
bootstrapping.

\stdpopsim is intended to be a fully open, community-developed project.  
Our implementations of genome representations and genetic maps for the three most
common study systems in computational genetics---humans, \emph{Drosophila},
and \emph{Arabidopsis}---are only intended to be a starting point for
future development.  
Researchers are invited to contribute to the resource by adding their
organisms and demographic models of choice. The \stdpopsim resource is
accompanied by clearly documented standard operating procedures that are
intended to minimize barriers to entry for new developers.  In this way, we
expect the resource to expand and adapt to meet the evolving needs of the
population genomics community.

@@

\pagebreak
\beginsupplement
\section*{Supplemental Figures}
\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_mask_Gutenkunst.pdf}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Estimates of population
size over time ($N(t)$) inferred using 4 different methods, \smcpp, \stairwayplot,
\MSMC with $n=2$ and $n=8$. Data were generated by simulating
replicate human genomes under the \cite{gutenkunst2009inferring} model and using the genetic map
inferred in \cite{international2007second}. From top to bottom we show estimates for each
of the three populations in the model: YRI, CEU, and CHB. In shades of blue we show the estimated
$N(t)$ trajectories for each replicate. In black we show the true population size history as inferred
for the rate of coalescence in the demographic model.}
\label{fig:n_t_gutenkunst}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{display_items/homo_sapiens_constant.pdf}
\caption{\textbf{Comparing estimates of $N(t)$ in humans}. Here we show estimates of population
size over time ($N(t)$) inferred using 4 different methods, \smcpp, \texttt{stairwayplot},
\MSMC with $n=2$ and $n=8$. Data were generated by simulating
replicate human genomes under a constant sized population model with $N=10^4$ and using the
HapMapII genetic map~\citep{international2007second}. In black we show the true population size history
of the model.}
\label{fig:n_t_HomSap_constant}
\end{center}
\end{figure}
